Article title

### 2017-07-27 paper-webserver.tex
DEL: unecessary words

% costs of bacterial infections
As food-sourcing is globalized, the consumer risk for bacterial infections is more complicated to track than ever before \cite{kozak2013foodborne}.
\textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common bacetrial infections originating from produce and is a central concern amongst public health efforts \cite{kozak2013foodborne}.
Foodborne illnesses translate into actual costs: in Canada alone, an estimated 4000 hospitalizations and 105 deaths per year are thought to be the result of various pathogens \cite{thomas2015estimates}.
Futhermore, the associated economic costs have been known for decades \cite{todd1989costs}.

% surveillance
The surveillance of bacterial pathogens is a routine function of public-health agencies in an effort to lower the burden caused by human illness.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, culture-dependent subtyping techniques are lately coupled with culture-independent methods based on whole-genome sequencing (WGS), to monitor known pathogens and detect emerging threats.

% traditional reference lab techniques
Subtyping encompasses various methods to designate particular strains of bacterial isolates.
In order to monitor outbreaks, platforms must rapidly subtype collected strains and communicate such information to other agencies \cite{ronholm2016navigating}. 
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, detecting the presence of specific cell surface antigens as tested by antibody reaction, pulsed-field gel electrophoresis (PFGE), to identify band patterns, known virulence factors (VFs), or antimicrobial resistance genes (AMR genes), or Sanger sequencing, to determine multi-locus sequence types (MLSTs) \cite{ronholm2016navigating}.
Using PFGE, outbreaks can be tracked by comparing band patterns of bacterial isolates, as routinely performed in public health agencies \cite{kozak2013foodborne}. 
The original computational methods to leverage WGS focused on creating \textit{in-silico} approaches to return subtyping results matching those of wet-lab techniques.
Software such as VirulenceFinder \cite{joensen2014real} compares a pre-collected set of reference sequences encompassing known VF gene sequences against a WGS result.
Similar approaches have been applied to predict AMR genes \cite{mcarthur2013comprehensive}, O-type, and H-type \cite{whiteside2016superphy}.

% potential of WGS
WGS has the potential to expand on traditional methods: due to the inclusion of the entire genome, WGS-based methods offer better resolution and context in evaluating the spread of outbreaks \cite{ronholm2016navigating}.
In the case for laboratory-based MLST, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, newer WGS-based methods such as core genome MLST (cgMLST) attempt to cover every common gene.
Secondary analyses, the study of relations and historical relevance of genomes, are now possible as a result of the large datasets provided by WGS \cite{ronholm2016navigating}.
Advances in WGS technologies have also enabled the monitoring and sharing of outbreak genomes in real-time \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-based methods still examine the underlying genetics, not the expressed phenotype; the presence of a particular gene sequence does not necessitate expression.
Correspondingly, a greater number of target sequences are identified using WGS approaches which complicates risk assessments \cite{ronholm2016navigating}, and laboratory based phenotypic tests remain relatively inexpensive \cite{koser2012routine}.
The effectiveness of WGS-based methods is also reliant on the accuracy and breadth of reference datasets.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc}, without employing software engineering principles \cite{de2015trends}.
Such designs were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
As WGS-technology expanded the availability of data, systems engineering principles became important as code failure could quickly disrupt entire bioinformatics pipelines \cite{schatz2015biological}.
The field also lags behind other big data industries: bioinformatics software typically struggles to perform computations in real-time, stores data in inflexible databases, doesn't support live visualizations \cite{de2015trends}, and lacks a common application programming interface (API) linking resources \cite{swaminathan2016review}.

% Solution via pre-computed analyses for routine tasks
Efforts such as the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} attempt to alleviate data concerns by curating a set of known AMR genes.
In addition, agencies such as the Food and Drug Administration (FDA), which hosts GenomeTrakr \cite{allard2016practical}, the National Center for Biotechnology Information (NCBI), which hosts GenBank \cite{doi:10.1093/nar/gks1195}, and the Center for Disease Control and Prevention (CDC), which hosts PulseNet \cite{swaminathan2001pulsenet}, provide services to share genome sequences and associated metadata either openly or amongst member institutions.
Recent efforts have been focused on providing modern tools for quality control of reference genomes \cite{parks2015checkm}, particularly amongst human-genome projects \cite{lee2017ngscheckmate}.
Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, and chewBBACA \url{https://github.com/mickaelsilva/chewBBACA} are available to compute core and accessory regions and identify SNPs across a large collection of reference genomes.
Taken together, open-access initiatives and shared tooling aim to alleviate the drawbacks of WGS-based methods by pre-computing sufficient reference data.

@@

% why Superphy
Subtyping and reference data generation often require command-line or programming experience.
To make such \textit{in-silico} techniques user-friendly for front-line clinicians and laboratories, a variety of web services have been developed in Denmark \cite{joensen2014real}, Taiwan \cite{liu2016construction}, and in collaborations \cite{hasman2015detection}.
While useful, arguably \cite{fricke2014bacterial}, the lack of bioinformatic standards and common infrastructure technologies have made it difficult to fully leverage the big-data results from WGS.

% Superphy
Previously we developed Superphy \cite{whiteside2016superphy}, an online platform targeting \textit{E. coli}, which determines Shiga toxin subtypes, AMR genes using CARD, and provides statistical significance tests between subgroups in associated metadata and pan-genome regions.
However, we struggled to provide real-time computing of results and to accomdate an ever-growing reference database.
Furthermore, due to the nature of the various dependencies involved, the platform was difficult for other researchers to independently replicate.

% this study
Here we present Spfy, a complete rewrite of Superphy which expands on supported features to also provide long-term storage of results in a graph database - one of the emerging \cite{de2015trends} database types being used for biological data.
We also containerize services using Docker, a modern approach found in many web servers and proposed to mitigate the complexities of bioinformatic pipelines \cite{di2015impact}.
Docker integration ensures that nested dependency resolution, which typically must be manually installed \cite{doi:10.1093/bioinformatics/btu153,laing2010pan,inouye2014srst2}, are handled automatically.
Furthermore, Spfy takes a modular approach to tool integration and distributes analyses over task queues, thereby allowing modules to run in parallel and address code failures.
These changes enable us to provide real-time genome analysis and database storage of results for upwards of 5000 genomes.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.

>>>>

% costs of bacterial infections
Consumers' risks for bacterial infections is more difficult to track due to the global sourcing of food \cite{kozak2013foodborne}.
Public health efforts monitor bacterial infections, of which \textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common \cite{kozak2013foodborne}.
Foodborne illnesses translate into actual costs: in Canada alone, an estimated 4000 hospitalizations and 105 deaths per year are thought to be the result of various pathogens \cite{thomas2015estimates}.
Futhermore, the associated economic costs have been known for decades \cite{todd1989costs}.
% surveillance
Public-health agencies routinely look for emerging threats, attempting to lower the burden caused by human illness.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, culture-dependent subtyping techniques are now coupled with culture-independent methods based on whole-genome sequencing (WGS).

% traditional reference lab techniques
Subtyping encompasses methods designating particular strains of bacterial isolates.
To monitor outbreaks, platforms must rapidly subtype collected strains and be shareable with other agencies \cite{ronholm2016navigating}. 
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, detecting the presence of specific cell surface antigens as tested by antibody reaction, pulsed-field gel electrophoresis (PFGE), to identify band patterns, known virulence factors (VFs), or antimicrobial resistance genes (AMR genes), or Sanger sequencing, to determine multi-locus sequence types (MLSTs) \cite{ronholm2016navigating}.
Using PFGE, outbreaks can be tracked by comparing band patterns of bacterial isolates, as routinely performed in public health agencies \cite{kozak2013foodborne}. 
The original computational methods leveraging WGS focused on returning subtyping results matching those of wet-lab techniques.
Software such as VirulenceFinder \cite{joensen2014real} compares a pre-collected set of reference VF gene sequences against a WGS result.
Similar approaches are applied to predict AMR genes \cite{mcarthur2013comprehensive}, O-type, and H-type \cite{whiteside2016superphy}.

% potential of WGS
WGS expands on traditional methods: due to the inclusion of the entire genome, WGS-based methods offer better resolution and context for evaluating the spread of outbreaks \cite{ronholm2016navigating}.
In the case of laboratory-based MLST, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, newer WGS-based methods, such as core genome MLST (cgMLST), attempt to cover every common gene.
Secondary analyses, the study of relations and historical relevance of genomes, are possible as a result of the large datasets provided by WGS \cite{ronholm2016navigating}.
Advances in WGS technologies have also enabled the real-time monitoring and sharing of outbreak genomes \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-based methods still examine the underlying genetics, not the expressed phenotype; the presence of a particular gene sequence does not necessitate expression.
Correspondingly, a greater number of target sequences are identified using WGS approaches which complicates risk assessments \cite{ronholm2016navigating}, and laboratory based phenotypic tests remain inexpensive \cite{koser2012routine}.
The effectiveness of WGS-based methods also rely on the accuracy and breadth of reference datasets.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc} without employing software engineering principles \cite{de2015trends}.
Tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
As WGS-technology expanded the availability of data, systems engineering principles became important since code failure could quickly disrupt entire bioinformatics pipelines \cite{schatz2015biological}.
The field also lags behind other big data industries: bioinformatics software still struggles to perform computations in real-time, stores data in inflexible databases, doesn't support live visualizations \cite{de2015trends}, and lacks a common application programming interface (API) linking resources \cite{swaminathan2016review}.

% Solution via pre-computed analyses for routine tasks
Efforts such as the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} attempt to alleviate data concerns by curating a set of known AMR genes.
In addition, agencies such as the Food and Drug Administration (FDA), which hosts GenomeTrakr \cite{allard2016practical}, the National Center for Biotechnology Information (NCBI), which hosts GenBank \cite{doi:10.1093/nar/gks1195}, and the Center for Disease Control and Prevention (CDC), which hosts PulseNet \cite{swaminathan2001pulsenet}, provide services to share genome sequences and associated metadata either openly or amongst member institutions.
Recent efforts have focused on providing modern tools for quality control of reference genomes \cite{parks2015checkm}, particularly amongst human-genome projects \cite{lee2017ngscheckmate}.
Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, and chewBBACA \url{https://github.com/mickaelsilva/chewBBACA} are freely available to compute core and accessory regions and identify SNPs across a large collection of reference genomes.
Taken together, open-access initiatives and shared tooling are lessening the drawbacks of WGS-based methods, by pre-computing sufficient reference data.

@@

% why Superphy
Subtyping and reference data generation often require command-line or programming experience.
To make such \textit{in-silico} techniques user-friendly for front-line clinicians and laboratories, a variety of web services have been developed in Denmark \cite{joensen2014real}, Taiwan \cite{liu2016construction}, and in collaborations \cite{hasman2015detection}.
While useful, arguably \cite{fricke2014bacterial}, the lack of bioinformatic standards and common infrastructure technologies make it difficult to fully leverage the big-data results from WGS.

% Superphy
Previously we developed Superphy \cite{whiteside2016superphy}, an online platform targeting \textit{E. coli} which determines Shiga toxin subtypes, AMR genes using CARD, and provides statistical significance tests between subgroups in associated metadata and pan-genome regions.
However, we struggled to provide real-time computing of results and to accomdate an ever-growing reference database.
Furthermore, due to the nature of the various dependencies involved, the platform was difficult for other researchers to independently replicate.

% this study
Here we present Spfy, a complete rewrite of Superphy expanding on supported features to also provide long-term storage of results in a graph database - one of the emerging \cite{de2015trends} database types being used for biological data.
We also containerize services using Docker, a modern approach found in many web servers and proposed to mitigate the complexities of bioinformatic pipelines \cite{di2015impact}.
Docker integration ensures that nested dependency resolution, which typically must be manually installed \cite{doi:10.1093/bioinformatics/btu153,laing2010pan,inouye2014srst2}, are handled automatically.
Furthermore, Spfy takes a modular approach to tool integration and distributes analyses over task queues, thereby allowing modules to run in parallel and address code failures.
These changes enable us to provide real-time genome analysis and database storage of results for upwards of 5000 genomes.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.
